Converting Population-Based Simulators to Probabilistic Programs

### 2019-05-16 paper/neurlps2019/main.tex
updated

Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way.
However, with the myriad of probabilistic programming systems (PPSs) and the complexities involved in running 
each individual system, such as specifying a model in the given language and $\backslash$ or running inference, which means it can be difficult for users to take advantage of the system. 
In addition to this, even though there exists a myriad of simulators across many scientific domains 
and within industry, none can be used within existing PPSs without having to re-implement the simulator in the given PPS, something that is not typically feasible, nor practical.  
However, recent work
by \cite{baydin2018efficient} demonstrated a pathway for how to turn a particular
type of simulator, written in a set of 9 languages, into a probabilistic program via the raw random
number calls.
 But, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs.
We, in this work, extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology modeling, multi-agent modeling, financial modeling and so forth. In addition to this, we enable simulators written in 13 different languages to be easily converted in probabilistic programs, which we algorithmically outline. We demonstrate our PPS on a both a toy population simulator and two important malaria population-based simulators called OpenMalaria and EMOD.

@@

>>>>

Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way.
However, with the myriad of probabilistic programming systems (PPSs) and the complexities involved in running 
each individual system, such as specifying a model in the given language, means that it can be difficult to re-implement the myriad of scientific and industrial simulators within an existing PPSs, something that is not typically feasible, nor practical.  
However, recent work
by \cite{baydin2018efficient} demonstrated a pathway for how to turn a particular
type of event-based simulators into a probabilistic program. 
% via the raw random number calls of the simulator.
 But, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs.
We, in this work, extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology modeling, multi-agent modeling and financial modeling. In addition to this, we enable simulators written in 13 different languages to be easily converted in probabilistic programs, which we algorithmically outline. We demonstrate inference performed with our PPS on a both a toy population simulator and two important malaria population-based simulators called OpenMalaria and EMOD.

@@

Probabilistic programming\cite{gordon2014probabilistic} is paradigm that enables users to write models in a programmatic way. 
The source code is then transformed into a program density, representing a joint density.
By having direct access to the source code enables Probabilistic Programming Systems~(PPS)
to analyse the model code in unique way, enabling posterior inference
to be performed in an automated way, which makes PPSs flexible and adaptive.
